---
title: "BLASE for Hematopoesis"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#library(blase)
library(fs)
library(Seurat)
library(dplyr)
library(slingshot)
library(tradeSeq)
library(BiocParallel)
library(scran)
library(scater)
library(viridis)
```

First configure R:
```{r, randomseed}
RNGversion("3.5.0")
SEED = 7
set.seed(SEED)
```

```{r, concurrency}
N_CORES = 4
if (ami::using_ci()) {
  N_CORES = 2
}
```

## Download Data
```{r}
options(timeout=300) #5 mins

root_dir = tools::R_user_dir("BLASE", "data")
article_dir = fs::path(root_dir, "hematopoeisis")
if (!dir.exists(article_dir)) {
  dir.create(article_dir, recursive = TRUE)
}

# https://www.nature.com/articles/s41586-024-07066-z#data-availability
# https://figshare.com/articles/dataset/Annotated_Seurat_objects/23290004?file=41046020
# Young1.All.T1.Seurat.RDS
path = fs::path(article_dir, "Young1.All.T1", ext="RDS")
if (!file.exists(path)) {
  download.file('https://figshare.com/ndownloader/files/41046020', path)
}else {
  print("Using cached")
}
```
## Load Single Cell
```{r}
seurat_obj = readRDS(path)
seurat_obj
seurat_obj@meta.data

DimPlot(seurat_obj, reduction = "wnn.umap", group.by = "STD.CellType", label=TRUE)
DimPlot(seurat_obj, reduction = "wnn.umap", group.by = "STD_Cat")
DimPlot(seurat_obj, reduction = "wnn.umap", group.by = "STD_Cat2")

```

```{r}
table(seurat_obj$STD.CellType)
```


```{r}
#expected_trajectory = c(
#    'Refined.HSC',
#    'HSC',
#    'MPP',
#    'CMP',
#    'MEP',
#    'EryP'
#    )

expected_trajectory = c(
    'Refined.HSC',
    'HSC', # Haematopoeitic Stem Cell
    'MPP', # multipotent progenitors
    'CMP', # common myeloid progenitor
    'GMP', # granulocyte–monocyte progenitors
    'MDP', # Monocyte-dendritic cell progenitor
    'Mono' # Monocytes
)

start_cell_type = expected_trajectory[1]
end_cell_type = expected_trajectory[length(expected_trajectory)]

seurat_obj$STD.CellType <- factor(seurat_obj$STD.CellType, levels=expected_trajectory)

s = subset(
  seurat_obj,
  STD.CellType %in% expected_trajectory
)
DefaultAssay(s) = 'RNA'
s[['ATAC']] = NULL
s[['SCT']] = NULL

s <- NormalizeData(s, verbose = FALSE)
s = subset(s, cells = sample(colnames(s), size=6000))
s <- FindVariableFeatures(s, selection.method = "vst", nfeatures = 1500, verbose = FALSE)
hvg = VariableFeatures(s)
VariableFeaturePlot(s)
```

```{r}
all.genes <- rownames(s)
s <- ScaleData(s, features = all.genes, verbose = FALSE)
s <- RunPCA(s, n_pcs=40, features = VariableFeatures(object = s), verbose = FALSE)
ElbowPlot(s, ndims = 40)
```

```{r}
N_PCS = 20
s <- RunUMAP(s, dims = 1:N_PCS, verbose=FALSE)
DimPlot(s, reduction = "umap", group.by = "STD.CellType", label = TRUE)
```

```{r}
sce = as.SingleCellExperiment(s)
rm(seurat_obj, s)
gc()

normcounts(sce) = exp(logcounts(sce))

plotUMAP(sce, color_by="STD.CellType")
```

## Diffusion Pseudotime
Destiny plots from:
https://broadinstitute.github.io/2019_scWorkshop/functional-pseudotime-analysis.html#diffusion-map-pseudotime
```{r}
library(destiny)
library(ggbeeswarm)

# Make a diffusion map.
dm <- DiffusionMap(sce, n_pcs=N_PCS)

# Plot diffusion component 1 vs diffusion component 2 (DC1 vs DC2). 
reducedDim(sce, "DiffusionMap") <- eigenvectors(dm)
plotDiffusionMap(sce, color="STD.CellType")

tmp <- data.frame(DC1 = eigenvectors(dm)[, 1],
                  DC2 = eigenvectors(dm)[, 2],
                  Timepoint = sce$STD.CellType)
ggplot(tmp, aes(x = DC1, y = DC2, colour = Timepoint)) +
    geom_point() + scale_color_viridis(discrete = TRUE) + 
    xlab("Diffusion component 1") + 
    ylab("Diffusion component 2") +
    theme_classic()

# Plot by dc 1
sce$slingshot = NULL
sce$pseudotime_diffusionmap <- rank(eigenvectors(dm)[,1])    # rank cells by their dpt
cd <- as.data.frame(colData(sce))

ggplot(as.data.frame(cd), 
       aes(x = pseudotime_diffusionmap, 
           y = STD.CellType, colour = STD.CellType)) +
    geom_quasirandom(groupOnX = FALSE) +
    scale_color_viridis(discrete = TRUE) + theme_classic() +
    xlab("Diffusion component 1 (DC1)") + ylab("Timepoint") +
    ggtitle("Cells ordered by DC1")

# ----------------------------------------------------------
sce$index = seq_len(ncol(sce))
start = sce$index[sce$index[sce$STD.CellType == start_cell_type]][1]
end = sce$index[sce$index[sce$STD.CellType == end_cell_type]][1]

dpt <- DPT(dm, tips = c(start, end))

sce$pseudotime_dpt <- rank(dpt$dpt) 
ggplot(as.data.frame(colData(sce)), 
       aes(x = pseudotime_dpt, 
           y = STD.CellType, colour = STD.CellType)) +
    geom_quasirandom(groupOnX = FALSE) +
    scale_color_viridis(discrete=TRUE) + theme_classic() +
    xlab("Diffusion map pseudotime (dpt)") +
    ylab("Timepoint") +
    ggtitle("Cells ordered by diffusion map pseudotime")

plotUMAP(sce, color_by="pseudotime_dpt")
plotDiffusionMap(sce, color="pseudotime_dpt")
# ----------------------------------------------------------
```
### Run Slingshot
```{r}
sce$slingPseudotime_1 = NULL
sce$slingPseudotime_2 = NULL
sce$slingshot = NULL

#sce <- slingshot(sce, reducedDim = 'DiffusionMap', clusterLabels="STD.CellType", start.clus=start_cell_type, end.clus=end_cell_type)
sce <- slingshot(sce, reducedDim = 'UMAP', clusterLabels="STD.CellType", start.clus=start_cell_type, end.clus=end_cell_type)
plotDiffusionMap(sce, color_by="slingPseudotime_1")
plotUMAP(sce, color_by="slingPseudotime_1")
```
```{r}
# subset out only highly variable genes for tradeseq
sce = sce[hvg,]
```


```{r}
# evaluateK(sce, k=5:9, parallel=T, BPPARAM=MulticoreParam(2), nGenes=length(rownames(sce)))
```

```{r}
sce <- fitGAM(sce, parallel=T, BPPARAM=MulticoreParam(2), nknots=9)
```

```{r}
associationTestResult <- associationTest(sce, lineages=T, global=F, contrastType="consecutive")
head(associationTestResult)
nrow(associationTestResult)
```
## Set up BLASE
```{r}
genelist = blase::get_top_n_genes(associationTestResult, n_genes = 1000, lineage = 1)
res = blase::find_best_params(sce, genelist, split_by="pseudotime_range", bins_count_range=c(10,12,14,16, 18),gene_count_range=c(10,20,40,80,160, 240, 320))

blase::plot_find_best_params_results(res)
```

```{r}
blase_data = as.BlaseData(sce, n_bins=10)
blase_data@genes = genelist[1:320]
evaluate_parameters(blase_data, make_plot = TRUE)
```

## Generate "bulk" datasets
In this article we will pseudobulk the Single Cell data into bulk datasets to demonstrate that
BLASE can map a given cluster back to itself.
```{r}
bulks_df = DataFrame(row.names = rownames(counts(sce)))
for (type in unique(sce$STD.CellType)) {
  bulks_df = cbind(bulks_df, rowSums(normcounts(subset(sce, , STD.CellType==type))))
}
colnames(bulks_df) = as.character(unique(sce$STD.CellType))
```

## Run BLASE
We will now verify that BLASE can accurately map these "bulk" datasets back to the SC clusters

```{r}
results = c()
for (type in expected_trajectory) {
  print(type)
  results = c(results, map_best_bin(blase_data, type, bulks_df))
}
```

```{r}
results
plot_mapping_result_heatmap(as.list(results))
```
