---
title: "Generating Weng 2024 Haematopoiesis Test Dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating Weng 2024 Haematopoiesis Test Dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, eval=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Generating Weng 2024 Haematopoiesis Test Dataset

This article, which is not run for this website, shows how we prepare the Haematopoeisis Dataset (a subset of the total data in the paper) ([Weng et al 2024](https://www.nature.com/articles/s41586-024-07066-z) 10x single cell data. It can be used to generate it from scratch, but we provide the processed object with BLASE.

```{r, setup, eval=FALSE}
library(readxl)
library(Seurat)
library(scran)
library(slingshot)
library(tradeSeq)
library(BiocParallel)
library(fs)
library(gridExtra)
library(scater)
library(ami)
library(phateR)
```

First configure R:
```{r, randomseed, eval=FALSE}
RNGversion("3.5.0")
SEED <- 7
set.seed(SEED)
```

```{r, concurrency, eval=FALSE}
N_CORES <- 4
if (ami::using_ci()) {
    N_CORES <- 2
}
```

First download the data from the Malaria Cell Atlas
```{r, getData, eval=FALSE}
options(timeout = 600) # 10 mins

root_dir <- tools::R_user_dir("BLASE", "data")
article_dir <- fs::path(root_dir, "generate-weng-2024-haematopoiesis-object")
article_dir
if (!dir.exists(article_dir)) {
    dir.create(article_dir, recursive = TRUE)
}

# https://figshare.com/articles/dataset/Annotated_Seurat_objects/23290004?file=41046020
# Young1.All.T1.Seurat.RDS
path <- fs::path(article_dir, "Young1.All.T1", ext = "RDS")
if (!file.exists(path)) {
    download.file("https://figshare.com/ndownloader/files/41046020", path)
} else {
    print("Using cached")
}
```

## Prepare SC
Now we read the files and create an SCE.

```{r, loadSC, eval=FALSE}
seurat_obj <- readRDS(path)

expected_trajectory <- c(
    "Refined.HSC",
    "HSC", # Haematopoeitic Stem Cell
    "MPP", # multipotent progenitors
    "CMP", # common myeloid progenitor
    "GMP", # granulocyte–monocyte progenitors
    "MDP" # Monocyte-dendritic cell progenitor
    # 'Mono' # Monocytes - excluded as unclear if part of true haematopoiesis process or mature and BM resident
)

start_cell_type <- expected_trajectory[1]
end_cell_type <- expected_trajectory[length(expected_trajectory)]

seurat_obj$STD.CellType <- factor(seurat_obj$STD.CellType, levels = expected_trajectory)

s <- subset(
    seurat_obj,
    STD.CellType %in% expected_trajectory
)
DefaultAssay(s) <- "RNA"
s[["ATAC"]] <- NULL
s[["SCT"]] <- NULL

s <- NormalizeData(s, verbose = FALSE)
s <- subset(s, cells = sample(colnames(s), size = 6000))
s <- FindVariableFeatures(s, selection.method = "vst", nfeatures = 1500, verbose = FALSE)
hvg <- VariableFeatures(s)
VariableFeaturePlot(s)
```

```{r, eval=FALSE}
all.genes <- rownames(s)
s <- ScaleData(s, features = all.genes, verbose = FALSE)
s <- RunPCA(s, n_pcs = 40, features = VariableFeatures(object = s), verbose = FALSE)
ElbowPlot(s, ndims = 40)
```

```{r, eval=FALSE}
N_PCS <- 20
s <- RunUMAP(s, dims = 1:N_PCS, verbose = FALSE)
```

## Create SCE
Now we can create a `SingleCellExperiment` object:
```{r, sceCreation, eval=FALSE}
sce <- as.SingleCellExperiment(s)
rm(seurat_obj, s)
gc()
normcounts(sce) <- exp(logcounts(sce))
plotUMAP(sce, color_by = "STD.CellType")
```

## Running Trajectory Analysis

Generate the pseudotime trajectory
```{r, eval=FALSE}
sce <- slingshot(sce, reducedDim = "UMAP", clusterLabels = "STD.CellType", start.clus = start_cell_type, end.clus = end_cell_type)
plotUMAP(sce, color_by = "slingPseudotime_1")
```

```{r, eval=FALSE}
# subset out only highly variable genes for tradeseq
sce <- sce[hvg, ]
```

```{r, eval=FALSE}
evaluateK(sce, k = 5:9, parallel = T, BPPARAM = MulticoreParam(2), nGenes = length(rownames(sce)))
```

```{r, eval=FALSE}
sce <- fitGAM(sce, parallel = T, BPPARAM = MulticoreParam(2), nknots = 9)
```

## Saving the SCE Object
Now we save the RDS so we can distribute with BLASE.
```{r, saveRDS, eval=FALSE}
processed_weng_2024_haematopoiesis <- sce
usethis::use_data(processed_weng_2024_haematopoiesis, overwrite = TRUE, compress = "bzip2")
```

## Loading the SCE Object
You can load this SCE with:
```{r, loadRDS}
data(processed_weng_2024_haematopoiesis, package = "blase")
processed_weng_2024_haematopoiesis
```

## Session Info
``` {r}
sessionInfo()
```
